---
title: "The Pink Ribbon"
author: "Jada Douthit & Patrick Stephens"
date: "2026-05-06"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```
## Background Information
This project uses a breast cancer biopsy dataset to study whether cell nucleus measurements can help identify malignant tumors. The dataset includes 569 biopsy records and 32 variables. The diagnosis variable classifies tumors as malignant (M) or benign (B). Other variables measure features such as radius, area, concavity, and texture of the cell nuclei.

Our goal was to determine which measurement is the strongest indicator of a malignant tumor by comparing malignant and benign cases using graphs and summary statistics.



## Research Question
### Which measurement is the most telltale sign of a malignant tumor?
#### Variables Used

- **Diagnosis**: categorical variable with two levels  
  - M = Malignant  
  - B = Benign  

- **Radius, concavity, concave points, and related nucleus measurements**: continuous variables  

- **Key comparison variables used in the graphs**:  
  - concave.points_worst  
  - concavity_worst  
  - radius_mean  
  
## Data Preparation
We loaded the dataset using read.csv() and checked the number of rows, columns, and variable names using nrow(), ncol(), and colnames(). We also created a frequency table for the diagnosis variable to compare the number of malignant and benign tumors.

Next, we calculated the percentage of each diagnosis group and created a pie chart to show the distribution of tumors. We checked for missing values using sum(is.na(BC)) to ensure the dataset was complete.

```{r starter}
library(dplyr)
library(tidyr)
library(tidyverse)
library(ggplot2)
BC = read.csv("/Volumes/T9/RSTUDIO WORK/Directed Seminar/Final-Project-DSS/breast-cancer.csv")
```

```{r inspect}
ncol(BC)
nrow(BC)
colnames(BC)
sum(is.na(BC))
table(BC$diagnosis)

MT <- BC |>
  filter(diagnosis == "M")


BT <- BC |>
  filter(diagnosis == "B")
```
  

## Graphical Analysis and Findings
### Diagnosis Distribution Pie Chart
The pie chart of diagnosis distribution shows the proportion of malignant versus benign tumors in the dataset. This graph gives a quick overview of how balanced the sample is and helps us understand the size of each group before making deeper comparisons.

```{r pie chart}
#Percent of M vs B
d_data <- BC |>
  count(diagnosis) |>
  mutate(
    percent = round((n / sum(n)) * 100,2)
  )

ggplot(d_data, aes(x = "", y = percent, fill = diagnosis)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +  # Convert to pie chart
  geom_text(aes(label = paste0(percent, "%")), 
            position = position_stack(vjust = 0.5), size = 3, color = "black") +
  labs(title = "Diagnosis Distribution") +
  theme_void() 
```  

### ECDF Plot of concave.points_worst
We created an ECDF graph for concave.points_worst by diagnosis. This graph is especially useful because it shows how quickly values build up in each group.

From the plot, malignant tumors appear shifted to the right compared to benign tumors. That means malignant tumors tend to have higher concave point values, while benign tumors tend to have lower values. This suggests that concave points are strongly associated with malignancy and may be one of the clearest indicators of a tumor’s seriousness.

```{r ecdf}
BC |> 
  filter(diagnosis %in% c("M", "B")) |>
  ggplot(aes(x = concave.points_worst, color = diagnosis)) +
  stat_ecdf(linewidth = 1) +
  theme(legend.position = "bottom")
```  

### Boxplot of concavity_worst by Diagnosis
The boxplot comparing concavity_worst across diagnosis groups also shows a clear difference between benign and malignant tumors. Malignant tumors tend to have higher concavity values and greater variability, while benign tumors are generally lower and more compact.

This pattern supports the idea that abnormal tumor shape and irregular cell boundaries are important signs of malignancy.

```{r BP}
boxplot(
  concavity_worst ~ diagnosis,
  data   = BC,
  main   = "Concavity (Worst) by Diagnosis",
  xlab   = "Diagnosis (B = Benign, M = Malignant)",
  ylab   = "Concavity Worst",
  col    = c("lightblue", "lightpink"),
  border = "gray40"
)
``` 

### Linear Comparison Idea
We also examined radius_mean in relation to concavity_mean to understand how overall tumor size relates to shape irregularity. Radius captures the average size of the cell nucleus, while concavity measures how indented or irregular the tumor boundary is. Since malignant tumors tend to be both larger and more irregular in shape, comparing these two variables helps reveal whether increases in size are associated with increases in structural irregularity across benign and malignant cases.

```{r LR}
BC |> 
  ggplot(aes(x = radius_mean, y = concavity_mean, color = diagnosis)) +
  geom_point(size = 2, alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 2) +
  ggthemes::scale_color_colorblind()
```
  
### Stacked Histogram of radius_worst by Diagnosis
We created a stacked histogram to compare radius_worst between malignant and benign tumors. The graph showed that malignant tumors were concentrated at higher radius values, while benign tumors were grouped more at lower values.

Although there was some overlap, the overall pattern suggests that malignant tumors tend to have larger radius measurements. This indicates that radius_worst is another useful feature for distinguishing between the two groups.

```{r hist}
BC |> 
  filter(diagnosis %in% c("M", "B")) |>
  ggplot(aes(x = radius_worst, fill = diagnosis)) +
  geom_histogram(bins = 15)
```


## Interpretation
The graphs suggest that measurements related to tumor shape irregularity, especially concave.points_worst and concavity_worst, are the strongest indicators of malignant tumors.

Malignant tumors generally have more irregular boundaries, higher concavity values, and larger extreme measurements. The “worst” measurements appear especially important because they capture the most abnormal cells in the biopsy sample.

## Conclusion
Our analysis found that concavity and concave point measurements, especially the worst values, are the clearest signs of malignant tumors in this dataset. Overall, this project shows how data analysis and mathematics can help support medical diagnosis by identifying patterns that separate malignant and benign tumors.


  
  
  
  
  
  
